---
title: "R Boxplots"
author: "Jesse Connell"
date: "4/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 10)
set.seed(0)
```

What is R's `boxplot` actually plotting?  For example, with a set of values
following a normal distribution:

```{r}
val_count <- 5000
val_mean <- 100
val_sd <- 100
values <- rnorm(n = val_count, mean = val_mean, sd = val_sd)
```

We see this in a histogram versus a boxplot:

```{r}
plotticks <- val_mean + seq(-400, 400, 100)
layout(mat = matrix(1:2, ncol = 1))
par(mar = c(2,2,2,2))
hist(values, breaks = 30,
     xlab = NULL, main = NULL,
     xlim = range(plotticks),
     xaxt = "n",
     yaxt = "n")
axis(side = 1, at = plotticks)
boxplot(values, horizontal = TRUE,
        ylim = range(plotticks),
        yaxt = "n")
axis(side = 1, at = plotticks)
```

What defines the edges of the box and the whiskers?

`boxplot.stats` explains the details.  Mostly.

```{r}
boxplot.stats(values)
```

The core part is the "stats" element.  `?boxplot.stats` says `stats` is:

> "a vector of length 5, containing the extreme of the lower whisker, the lower
> 'hinge', the median, the upper 'hinge' and the extreme of the upper whisker."

## Hinges

> "The two 'hinges' are versions of the first and third quartile, i.e., close to
> `quantile(x, c(1,3)/4)`."

```{r}
quantile(values)
hinges <- unname(quantile(values, c(1,3)/4))
hinges
```

## Whiskers

> "If `coef` is positive, the whiskers extend to the most extreme data point
> which is no more than coef times the length of the box away from the box."

...what?  Well, with a default of 1.5 (see the `range` argument to `boxplot`)
and "the box" being defined as the "the hinges" elsewhere in the page, the
whiskers end up at the closest point within `+/- 1.5 * IQR` of the hinges, where
the interquartile range is the difference between the hinges.

```{r}
coeff <- 1.5
boxedges <- quantile(values, c(1,3)/4)
boxlen <- diff(boxedges)
whisker_bounds <- boxedges + c(-1, 1)*coeff*boxlen
# That was just the abstract limits for the whiskers.  The actual values depend
# on where our points are.
in_bounds <- whisker_bounds[1] < values & values < whisker_bounds[2]
whiskers <- range(values[in_bounds])
whiskers
```

## All Together

```{r}
boxplot.stats(values)$stats
sort(c(hinges, whiskers, median(values)))
```
